A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-01-22 ManuscriptGeophysics2018.tex
Corrections and additions re z_skin

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of 1e-12 and 1e-30, respectively. The
relative error is shown in (b), where the QWE result was taken as \emph{truth}.
For offsets greater than roughly 15.5\,km the relative error becomes
meaningless, as the QWE fails itself; this part is greyed out in the figure.

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
(1e-2) in all cases. The exception are large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz. In the shallow water case this
zone has a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
1e-25~V/m.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of 1e-25~V/m, and therefore far below the noise level.}
%

Another error test is shown in Figure~\ref{fig:skindepth1}. It shows in (a) the
amplitude of the analytical fullspace solution for a wide range of
resistivities and frequencies: $\rho=$ 1e-10$\,\Omega\,$m to 1e10\,$\Omega\,$m,
and $f =$ 1e-10\,Hz to 1e14\,Hz. In this case, the horizontal offset is set to
the skin depth, and the vertical source-receiver separation to a tenth of it,
where the skin depth is given by \citep[e.g., ][ equation~1.49]{B.SEG.88.Ward}
%
\begin{equation}
  z_\mr{skin} = \left[\omega \sqrt{\frac{\mu\varepsilon}{2}}
  \sqrt{\left\{1 + \left(\frac{1}{\omega\varepsilon\rho}\right\}^2 \right) -1}
  \right]^{-1} \, . \\
  \label{eq:skin}
\end{equation}
%
The contour lines in the figure show the magnitude of the offset: the line with
0 is where the offset is 1\,m, the line with 3 where the offset is 1\,km, etc.
The contour lines of the offset have a knee, after which they become almost
independent of frequency. This is where the quasi static approximation breaks
down, and wave phenomena become dominant. Subplot (b) shows the relative error
of the DLF using the filter \emph{Wer201} compared to the analytical solution.
It shows nicely that the filter can be used for all cases that are in the quasi
static region, but not outside.
%
\plot*{skindepth1}{width=.8\textwidth}{(a) Amplitude of the analytical
  fullspace solution for a wide range of resistivities and frequencies. The
  contour lines show the power of the offset used for the calculation, where
  the offset is set to the skin depth. (b) Relative error of the DLF compared
  with the analytical solution. The filter \emph{Wer201} works well in the
  diffusive limit, but fails when wave phenomena become dominant.}
%

>>>>

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of $10^{-12}$ and $10^{-30}$,
respectively. The relative error is shown in (b), where the QWE result was
taken as \emph{truth}. For offsets greater than roughly 15.5\,km the relative
error becomes meaningless, as the QWE fails itself; this part is greyed out in
the figure.

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
($10^{-2}$) in all cases. The exception are large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz. In the shallow water case this
zone has a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
$10^{-25}\,$V/m.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore far below the noise level.}
%

Another error test is shown in Figure~\ref{fig:skindepth1}. It shows in (a) the
amplitude of the analytical fullspace solution for a wide range of
resistivities and frequencies: $\rho=$ $10^{-10}\,\Omega\,$m to
$10^{10}\,\Omega\,$m, and $f =10^{-10}$\,Hz to $10^{14}$\,Hz. In this case, the
horizontal offset is set to the skin depth, and the vertical source-receiver
separation to a tenth of it, where the skin depth is given by \citep[e.g., ][
equation~1.49]{B.SEG.88.Ward}
%
\begin{equation}
  z_\mr{skin} = \left\{\frac{\omega^2 \varepsilon\mu}{2}
    \left[ \sqrt{1 + \frac{1}{\left(\omega\varepsilon\rho\right)^2}} -1\right]
  \right\}^{-1/2} \, . \\
  \label{eq:skin}
\end{equation}
%
The grey contour lines in the figure show the magnitude of the offset: the line
with 0 is where the offset is 1\,m, the line with 3 where the offset is 1\,km,
etc. The contour lines of the offset have a knee, after which they become
almost independent of frequency. This is where the quasi static approximation
breaks down, and wave phenomena become dominant. The red line indicates where
$\omega\varepsilon\rho = 1$. Subplot (b) shows the relative error of the DLF
using the filter \emph{Wer201} compared to the analytical solution. It shows
nicely that the filter can be used for all cases that are in the quasi static
region, but not outside.
%
\plot*{skindepth1}{width=.8\textwidth}{(a) Amplitude of the analytical
  fullspace solution for a wide range of resistivities and frequencies. The
  contour lines show the power of the offset used for the calculation, where
  the offset is set to the skin depth. (b) Relative error of the DLF compared
  with the analytical solution. The filter \emph{Wer201} works well in the
  diffusive limit, but fails when wave phenomena become dominant. The red
 line shows where $\omega\varepsilon\rho = 1$.}
%
